Modelling RNA-based oscillatory circuits

### 2015-09-16 MScWriteup.tex
final day!

%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Computational design of RNA-based oscillatory circuits}

@@

\begin{abstract}
For the synthetic biologist looking to engineer regulation of genetic circuitry, RNA is an attractive tool due to the ease of predicting its behaviour from physiochemical models. In this report, we introduce a system of ODEs to model a recently designed synthetic regulatory circuit, before attempting to estimate its unknown parameters using recent single cell time series data. We provide estimates for a subset of parameters, but find many of them to be inestimable. We discuss why this is the case for our system, and suggest how this problem might be resolved.
\end{abstract}

@@

As well as acting as the intermediate between DNA and protein, RNA molecules play direct and important roles in regulating gene expression \cite{Isaacs2006}. For the synthetic biologist looking to engineer regulation of genetic circuitry, RNA offers an attractive alternative to more traditional methods, which typically involve using proteins to regulate DNA transcription. In comparison to proteins, it is relatively straightforward to predict the intra and intermolecular interactions of an RNA from its sequence using physiochemical models. Recently, this has been exploited to computationally design synthetic sRNAs - small RNAs which do not code for a protein, but rather have some direct regulatory function -  with regulatory behaviour that can be predicted \cite{Rodrigo2013}\cite{Rodrigo2012}.

@@

\subsection{The sRNA regulatory system}
\label{The sRNA regulatory system}
In bacteria, one mechanism by which gene expression is naturally regulated is as follows \cite{Soper2010}: In order for a bacterial mRNA to be translated into a protein, the Ribosome must initially bind to the mRNA (Fig. \ref{RBS}). This occurs at the Ribosome Binding Site (RBS) \cite{Shine1974}, a specific nucleotide sequence found on the mRNA. 
\setcounter{figure}{1}
\begin{figure*}[t!]
\centering
\includegraphics[trim = 30 600 30 0,page=4,clip = true]{pnas1203831109.pdf}
\caption{A logical AND gate formed from a self repressed mRNA, and an sRNA which uncovers its RBS. In this system, transcription of the sRNA (transRNA) and mRNA (5'-UTR,GFP) are controlled by two promoters, $P_{\mathrm{LtetO-1}}$ and $P_\mathrm{LlacO-1}$. These are disabled by the presence of two chemical repressors, TetR and LacI, found naturally in the strain of \textit{E. coli} discussed. These chemical repressors are themselves disabled by two chemicals, aTC and IPTG. In the notation of the diagram, a barred line indicates repression, and an arrowed line indicates production. We see a `double negative' in aTc repressing TetR, which itself represses transcription of the sRNA (likewise for IPTG and the mRNA). Thus the presence of the sRNA and mRNA are controlled by the presence of aTc and IPTG, which can be experimentally introduced to the cell.  Image reproduced from \cite{Rodrigo2012}.}
\label{ANDGate}
\end{figure*}
\setcounter{figure}{0}

@@

In an mRNA there is an untranslated region of nucleotides at the 5' end of the molecule (the 5' UTR), upstream of the RBS \footnote{Both DNA and RNA are directional - the backbone of the molecule is not symmetric. This gives the molecule two distinct ends, denoted 5' and 3'. Translation can only occur in the 5' to 3' direction, hence the meaning of `upstream'}. Translation may be self repressed by this `tail' folding over and binding across the RBS, forming a  loop in the mRNA and preventing the Ribosome from binding (Fig. \ref{RBS}). This self repression may be released with an sRNA which binds to this looped tail - the new conformation of the sRNA:mRNA complex uncovers the RBS, allowing the Ribosome to bind. In summary, the presence of the sRNA positively regulates gene expresssion.

@@

In \eqref{eq:s} - \eqref{eq:z}, we present a modified version an existing model which describes the system \cite{Jaramilloa}, consisting of a set of ODE's with mass action kinetics \cite{UriAlon}.  Its state is given by the vector $(s,m,s:m,c,p,g,z)$, with all other variables representing model parameters. Tables \ref{StateVariables} and \ref{ModelParameters} give complete descriptions of the parameters and state variables.

@@

\begin{align}
\frac{\mathrm{d}\mathbf{y}}{\mathrm{d}t} = \mathbf{f}(\mathbf{y},\boldsymbol{\theta},t),
\end{align}
where $\mathbf{y}$ is our state vector, $\boldsymbol{\theta}$ is a vector of model parameters, and $t$ is time. The model may be integrated numerically, giving a prediction $\mathbf{y}(t,\boldsymbol{\theta})$. The least squares error between the model prediction and an experimental time series is defined as
\begin{align}
 \sum_{i =1}^{n} (\mathbf{y}_{\mathrm{exp}}(t_{i}) - \mathbf{y}(t_{i},\boldsymbol{\theta}))^2,
\end{align}
where the experimental time series, $\mathbf{y}_{\mathrm{exp}}(t_{i})$ is recorded at timepoints $t_{i}$, $i = 1 \hdots n$. This error function defines a landscape in $\boldsymbol{\theta}$ space, and we seek to minimise it by varying $\boldsymbol{\theta}$. In our case, we do not have experimental data on the full state vector, but only one component of it - the observed fluorescence, $z(t)$. In addition, rather than a single experimental run, we have many, corresponding to a time series from each cell. We incorporate this by fitting to the experimental mean of the data, and only minimising over the observed component. Our minimisation problem is thus
\begin{align}
\argmin_{\boldsymbol{\theta}} \sum_{i =1}^{n} (z_{\mathrm{exp, mean}}(t_{i}) - z(t_{i},\boldsymbol{\theta}))^2.
\label{eq:min}
\end{align}
The next step is performing the minimisation. In general, the landscape defined by the error function may be rugged and contain many local minima, which a local optimisation algorithm may get stuck in. To try and surmount this problem, \cite{Algorithms2003} suggests the use of a global optimisation algorithm, and in particular demonstrates that several Evolutionary Algorithms perform well on a test problem involving a set of ODEs modelling a biological system. We choose one of those recommended, the CMA-ES \cite{Hansen2006,Hansen2011}, which in addition. 
\todo[inline]{why use CMA-ES? Some refs}

@@

\begin{figure*}
    \begin{subfigure}[c]{0.49\textwidth}
    \centering
    \includegraphics[scale = 0.24,clip = true, trim = 115 0 60 0]{13_9_bestPlot}
        \caption{Model prediction, using the parameter set with the smallest error value of the initial 200 found, for the \textit{13\_9} dataset.}
                \label{InitialResults_13_9_best}
    \end{subfigure}
    \begin{subfigure}[c]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.24,clip = true, trim = 100 0 70 0]{14_7_bestPlot}
        \caption{Model prediction, using the parameter set with the smallest error value of the initial 200 found, for the \textit{14\_7} dataset.}
        \label{InitialResults_14_7_best}
    \end{subfigure}
        \begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.24, clip = true, trim = 140 0 125 0]{13_9_hist}
        \caption{Histogram of estimated parameter values, found from 200 runs of the CMA-ES algorithm. Fitted to the \textit{13\_9} dataset.}
        \label{InitialResults_13_9_hist}
    \end{subfigure}
    \begin{subfigure}[c]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.24,clip = true, trim = 140 0 125 0]{14_7_hist}
        \caption{Histogram of estimated parameter values, found from 200 runs of the CMA-ES algorithm. Fitted to the \textit{14\_7} dataset. }
        \label{InitialResults_14_7_hist}
    \end{subfigure}
    \caption{Parameter estimates and model predictions for the \textit{13\_9} and \textit{14\_7} datasets. Note that in the model predictions, aTc forcing is shown - IPTG concentration is constant at a level which saturates the cell's response. The forcing curve's height is schematic - aTc concentration is switched between off and on (where by `on' we mean a level which saturates the cell's response).}
\label{InitialResults}
\end{figure*}

@@

Figs. \ref{InitialResults_13_9_correlation}, \ref{InitialResults_14_7_correlation} show correlation matrices between parameter sets\footnote{These are computed by viewing each of the estimated parameter sets as a sample from a multivariate random variable. An element of the correlation matrix $R_{ij}$ is then the Pearson correlation coefficient between the $i^\mathrm{th}$ and $j^\mathrm{th}$ random variable - or, in our case, the $i^\mathrm{th}$ and $j^\mathrm{th}$ parameters}. These correlation matrices indicate that there are spaces of parameters within which the fitness function remains approximately constant - for example, in both datasets there exists a positive correlation between values of $\mu$ and $\beta$. Referring to \eqref{eq:p}, this makes heuristic sense - the two parameters may have similar effects on model predictions, and may be able to co-vary in such a way as to leave the model prediction unchanged.
The results suggest a fitness landscape relatively flat to perturbations in certain combinations of parameters, and indicate we may have difficulty obtaining unique estimates of the model parameters.

@@

@@

 Taken together, these results suggest that simply performing a least squares fit on the available data will not give unique estimates for many of the parameters. In the following sections we investigate this further.

@@

We see that many of the parameters give sensitivity curves of similar shapes, and have near linear dependence - this implies that the effects of perturbing any one of these parameters all look similar in terms of model output, and are hard to distinguish between. This may help to explain why some of our initial parameter estimates are very loose, and why there are correlations between the estimated parameter sets - this result suggests there is a family of parameters all of which, in terms of the model output we have available, cannot be resolved. As such, we should view estimates of these parameters with extreme caution. Note that the sensitivity curve that looks least similar to the others in Fig. \ref{SensitivityMatrix_scaled} - $\mu$ - corresponds to a relatively tightly estimated value in Fig. \ref{InitialResults}. Note also that the curve for $\Theta$ is very similar to those for the remaining parameters. This suggests that our estimated value of $\Theta$ in Fig. \ref{InitialResults_13_9_hist} is misleading, and likely a consequence of the initial bounding box.

@@

Fig. \ref{model_output} shows model output for all state variables, using the parameter values giving the lowest error on the \textit{13\_9} dataset, and normalised to lie on the same scale. We see that $s,m,s:m$ and $c$ respond rapidly to the forcing function, flipping between the two fixed points defined by the step function forcing almost instantly. By contrast, there is delay in the response of $p$ and $g$, on a timescale comparable to the period of the forcing. This result suggests that, at least in some of the parameter sets initially estimated, the system may have two timescales in it - a fast timescale in which \eqref{eq:s} - \eqref{eq:c}, representing the hybridization of the sRNA and mRNA into a stable complex, equilibrate in response to external forcing, and a slower timescale, in \eqref{eq:p}, \eqref{eq:g}, in which measured fluorescence changes in response to the forcing. Biologically, this would correspond to \eqref{eq:p}, representing translation, being a rate limiting step.

@@

This would explain the similarity between the sensitivity curves of many parameters - if the parameters in \eqref{eq:s} - \eqref{eq:c} can only act to alter the fixed point values of $p$ and $g$ which the system tends toward, they all play qualitatively identical roles on the time series of $z$. The fact that $\beta m +f_{s}\beta c$ is the only term in \eqref{eq:p}, \eqref{eq:g} to include $f_{s}$ and $\beta$ may also explain the similarity of the sensitivity curves for these parameters to those found in \eqref{eq:s} - \eqref{eq:c} (Fig. \ref{SensitivityMatrix_scaled}), by the same reasoning. This explanation also suggests that all parameter sets giving the same value of the model's fixed point will give similar model predictions - and there may be many, very different, parameter sets which all give the same fixed point. Fig. \ref{fixedpoint_f} shows a scatterplot of error function value against $\beta m +f_{s}\beta c$ for the 200 estimated parameter sets for the \textit{13\_9} dataset, and demonstrates that, although individual parameter values can be spread across very large ranges, they are correlated in such a way as to give similar model fixed points.

@@

 In contrast to the results of Fig. \ref{InitialResults}, we see relatively tightly estimated parameters, with no estimated values near the initial bounding box. Further, this simplified model achieves error values as low as the more complex model. Fig \ref{13_9_f_simplified} shows that the simplified algorithm is more consistently able to find the minimal function value identified in Fig. \ref{InitialResults_13_9_f}, possibly because the low dimension of the new model makes parameter space easier to search.

 We note that \ref{13_9_scatterplot_simplified} shows a strong positive correlation between the minima found for $F$ and $\Theta$. This is consistent with the  normalised sensitivity curves for the \textit{13\_9} dataset, shown in Fig. \ref{Sensitivity_scaled_simplified}. As before, the similarity of the $F$ and $\Theta$ curves suggests that any minima in $F - \Theta$ space will be locally non-unique \cite{Beck}. This may be explained as follows: neglecting the ClpX degradation term, \eqref{eq:p_simplified} is a simple exponential decay to a fixed value. Here, $F$ will determine the value the $p$ will tend to, and $\mu$ will determine its rate constant. $F$ is then a linear scaling factor, just as $\Theta$ is - both scale the value the exponential will tend to.

@@

 We also note that the parameter estimates come in two distinct clusters.  We further investigate by fixing the value of $\Theta$, and plotting the error as a function of $\mu$ and $F$. Results are shown in Fig. \ref{SimplifiedModelLikelihood}. We find a local basin in $ \mu - F$ space, pocketed by many local minima of similar fitness values. The optimisation algorithm will always terminate in one of these local minima, explaining the disconnected clusters shown in Fig.  \ref{13_9_scatterplot_simplified}. Thus the tightness of each peak in Fig. \ref{13_9_hist_simplified} is misleading - the uncertainty in parameter estimates is better reflected by the contours of Fig. \ref{Likelihood_profile_fine}.

@@

In this report, we have presented a system of ODE's to model a recently proposed synthetic RNA regulatory circuit \cite{Rodrigo2012}, and attempted to estimate the model's parameters using existing time series data, with a least squares minimisation approach similar to that found in recent synthetic biology literature \cite{Hu2015}. We have found many of the parameters contained in the full model to be inestimable, and suggested reasons for why this might be. We have further suggested a simplified model, which provides an equally good description of the data with fewer parameters than the initial model.

@@

On the other hand, if a good description of the fluorescence data is what is really of interest, the simplified model may be a better starting point. On the experimental side, $\mu$ and $\Theta$ are both parameters which can be easily estimated independent of our datasets - a realistic range of values could be used to bound the error function profile shown in Fig. \ref{SimplifiedModelLikelihood}. Further modelling work may focus on improving the modelling of translation, the rate limiting step - \cite{Hu2015}, for example, uses a more detailed model than our single step description of translation. Any additional parameters introduced will likely act on the same timescale as $\mu$, and stand a better chance of being estimable.

@@

% references section
\begin{footnotesize}
\bibliographystyle{./IEEE_bibtex_class/IEEEtran}
\bibliography{./IEEE_bibtex_class/IEEEabrv,library}
\end{footnotesize}
\clearpage
\onecolumn

@@

>>>>

%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{Modelling RNA-based oscillatory circuits}

@@

\begin{abstract}
For the synthetic biologist looking to engineer regulation of genetic circuitry, RNA is an attractive tool due to the ease of predicting its behaviour from physiochemical models. In this report, we introduce a system of ODEs to model a recently designed synthetic regulatory circuit, before attempting to estimate its unknown parameters using recent single cell time series data. We provide estimates for a subset of parameters, but find many of them to be inestimable. We suggest this is a consequence of translation being a rate limiting step in our ODE model and suggest that a simplified model, describing translation alone, may describe the data equally well.
\end{abstract}

@@

As well as acting as the intermediate between DNA and protein, RNA molecules play direct and important roles in regulating gene expression \cite{Isaacs2006}. For the synthetic biologist looking to engineer regulation of genetic circuitry, RNA offers an attractive alternative to more traditional methods, which typically involve using proteins to regulate DNA transcription. In comparison to proteins, it is relatively straightforward to predict the intra and intermolecular interactions of RNA from its sequence using physiochemical models. Recently, this has been exploited to computationally design synthetic sRNAs - small RNAs which do not code for a protein, but rather have some direct regulatory function -  with regulatory behaviour that can be predicted \cite{Rodrigo2013}\cite{Rodrigo2012}.

@@

\subsection{The sRNA regulatory system}
\label{The sRNA regulatory system}
In bacteria, one mechanism by which gene expression is naturally regulated is as follows \cite{Soper2010}: In order for a bacterial mRNA to be translated into a protein, the Ribosome must initially bind to the mRNA (Fig. \ref{RBS}). This occurs at the Ribosome Binding Site (RBS) \cite{MolecularBiology}, a specific nucleotide sequence found on the mRNA. 
\setcounter{figure}{1}
\begin{figure*}[t!]
\centering
\includegraphics[trim = 30 600 30 0,page=4,clip = true]{pnas1203831109.pdf}
\caption{A logical AND gate formed from a self repressed mRNA, and an sRNA which uncovers its RBS. In this system, transcription of the sRNA (transRNA) and mRNA (5'-UTR,GFP) are controlled by two promoters, $P_{\mathrm{LtetO-1}}$ and $P_\mathrm{LlacO-1}$. These are disabled by the presence of two chemical repressors, TetR and LacI, found naturally in the strain of \textit{E. coli} discussed. These chemical repressors are themselves disabled by two chemicals, aTC and IPTG. In the notation of the diagram, a barred line indicates repression, and an arrowed line indicates production. We see a `double negative' in aTc repressing TetR, which itself represses transcription of the sRNA (likewise for IPTG and the mRNA). Thus the presence of the sRNA and mRNA are controlled by the presence of aTc and IPTG, which can be experimentally introduced to the cell.  Image reproduced from \cite{Rodrigo2012}.}
\label{ANDGate}
\end{figure*}
\setcounter{figure}{0}

@@

In an mRNA there is an untranslated region of nucleotides at the 5' end of the molecule (the 5' UTR), upstream of the RBS \footnote{Both DNA and RNA are directional - the backbone of the molecule is not symmetric. This gives the molecule two distinct ends, denoted 5' and 3'. Translation can only occur in the 5' to 3' direction.}. Translation may be self repressed by this `tail' folding over and binding across the RBS, forming a  loop in the mRNA and preventing the Ribosome from binding (Fig. \ref{RBS}). This self repression may be released with an sRNA which binds to this looped tail - the new conformation of the sRNA:mRNA complex uncovers the RBS, allowing the Ribosome to bind. In summary, the presence of the sRNA positively regulates gene expresssion.

@@

In \eqref{eq:s} - \eqref{eq:z}, we present a modified version an existing model which describes the system \cite{Jaramilloa}, consisting of a set of ODE's with mass action kinetics.  Its state is given by the vector $(s,m,s:m,c,p,g,z)$, with all other variables representing model parameters. Tables \ref{StateVariables} and \ref{ModelParameters} give complete descriptions of the parameters and state variables.

@@

\begin{align}
\frac{\mathrm{d}\mathbf{y}}{\mathrm{d}t} = \mathbf{f}(\mathbf{y},\boldsymbol{\theta},t),
\end{align}
where $\mathbf{y}$ is our state vector, $\boldsymbol{\theta}$ is a vector of model parameters, and $t$ is time. The model may be integrated numerically, giving a prediction $\mathbf{y}(t,\boldsymbol{\theta})$. The least squares error between the model prediction and an experimental time series is defined as
\begin{align}
 \sum_{i =1}^{n} (\mathbf{y}_{\mathrm{exp}}(t_{i}) - \mathbf{y}(t_{i},\boldsymbol{\theta}))^2,
\end{align}
where the experimental time series, $\mathbf{y}_{\mathrm{exp}}(t_{i})$ is recorded at timepoints $t_{i}$, $i = 1 \hdots n$. This error function defines a landscape in $\boldsymbol{\theta}$ space, and we seek to minimise it by varying $\boldsymbol{\theta}$. In our case, we do not have experimental data on the full state vector, but only one component of it - the observed fluorescence, $z(t)$. In addition, rather than a single experimental run, we have many, corresponding to a time series from each cell. We incorporate this by fitting to the experimental mean of the data, and only minimising over the observed component. Our minimisation problem is thus
\begin{align}
\argmin_{\boldsymbol{\theta}} \sum_{i =1}^{n} (z_{\mathrm{exp, mean}}(t_{i}) - z(t_{i},\boldsymbol{\theta}))^2.
\label{eq:min}
\end{align}
The next step is performing the minimisation. In general, the landscape defined by the error function may be rugged and contain many local minima, which a simple local optimisation algorithm may get stuck in. To try and surmount this problem, \cite{Algorithms2003} suggests the use of a global optimisation algorithm, and in particular demonstrates that several Evolutionary Algorithms perform well on a test problem involving a set of ODEs modelling a biological system. We choose one of those recommended, the CMA-ES \cite{Hansen2006,Hansen2011} which has been successfully used for parameter estimation in ODE models \cite{Ballesta2011,Dulong2015} .

@@

\begin{figure*}
    \begin{subfigure}[c]{0.49\textwidth}
    \centering
    \includegraphics[scale = 0.24,clip = true, trim = 115 0 60 0]{13_9_bestPlot}
        \caption{Model prediction, using the parameter set with the smallest error value of the initial 200 found, for the \textit{13\_9} dataset.}
                \label{InitialResults_13_9_best}
    \end{subfigure}
    \begin{subfigure}[c]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.24,clip = true, trim = 100 0 70 0]{14_7_bestPlot}
        \caption{Model prediction, using the parameter set with the smallest error value of the initial 200 found, for the \textit{14\_7} dataset.}
        \label{InitialResults_14_7_best}
    \end{subfigure}
        \begin{subfigure}[h]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.24, clip = true, trim = 130 0 125 0]{13_9_hist}
        \caption{Histogram of estimated parameter values, found from 200 runs of the CMA-ES algorithm. Fitted to the \textit{13\_9} dataset.}
        \label{InitialResults_13_9_hist}
    \end{subfigure}
    \begin{subfigure}[c]{0.49\textwidth}
    \centering
        \includegraphics[scale = 0.24,clip = true, trim = 130 0 125 0]{14_7_hist}
        \caption{Histogram of estimated parameter values, found from 200 runs of the CMA-ES algorithm. Fitted to the \textit{14\_7} dataset. }
        \label{InitialResults_14_7_hist}
    \end{subfigure}
    \caption{Parameter estimates and model predictions for the \textit{13\_9} and \textit{14\_7} datasets. Note that in the model predictions, aTc forcing is shown - IPTG concentration is constant at a level which saturates the cell's response. The forcing curve's height is schematic - aTc concentration is switched between off and on (where by `on' we mean a level which saturates the cell's response).}
\label{InitialResults}
\end{figure*}

@@

@@

Some aspects of these initial fitting results are worth comment. For example, Fig. \ref{InitialResults_13_9_f} shows that many of the estimated parameter sets in Fig. \ref{InitialResults_13_9_hist} do not give the minimum error function value found - instead, the majority find a local minimum slightly above it. In fact, if we examine the estimated parameter sets giving the lowest error values, several have $\mu = 0.05$, the upper bound set. Thus, in the tightly clustered $\mu$ histogram, we must bear in mind that $\mu \approx 0.044$ is not a unique optimum - the error landscape may indeed be pocketed by many local minima, all giving a similar quality of fit.

Taken together, these results suggest the following:

\begin{enumerate}
\item Simply performing a least squares fit on the available data will not give tightly constrained estimates for many of the parameters.
\item $\mu$ appears to be an important parameter in determining the fit.
\item Our initial bounding box may have been set incorrectly for several parameters - $\mu$, $\Theta$, for example - and further minima may lie outside it. 
\end{enumerate}

In the next section we further investigate the loose estimates for many of the parameters.

@@

@@

We see that many of the parameters give sensitivity curves of similar shapes, and have near linear dependence - this implies that the effects of perturbing any one of these parameters all look similar in terms of model output, and are hard to distinguish between. This may help to explain why some of our initial parameter estimates are very loose - this result suggests there is a family of parameters all of which, in terms of the model output we have available, cannot be resolved. As such, we should view estimates of these parameters with extreme caution. Note that the sensitivity curve that looks least similar to the others in Fig. \ref{SensitivityMatrix_scaled} - $\mu$ - corresponds to a relatively tightly estimated value in Fig. \ref{InitialResults}. Note also that the curve for $\Theta$ is very similar to those for the remaining parameters. This suggests that our estimated value of $\Theta$ in Fig. \ref{InitialResults_13_9_hist} is misleading, and likely a consequence of the initial bounding box.

@@

Fig. \ref{model_output} shows model output for all state variables, normalised to lie on the same scale  - it was made using the parameter values giving the lowest error on the \textit{13\_9} dataset, but the model output is very similar across all estimated parameter sets.
  We see that $s,m,s:m$ and $c$ respond rapidly to the forcing function, flipping between the two fixed points defined by the step function forcing almost instantly. By contrast, there is delay in the response of $p$ and $g$, on a timescale comparable to the period of the forcing. This result suggests that the system may have two timescales in it - a fast timescale in which \eqref{eq:s} - \eqref{eq:c}, representing the hybridization of the sRNA and mRNA into a stable complex, equilibrate in response to external forcing, and a slower timescale, in \eqref{eq:p}, \eqref{eq:g}, in which measured fluorescence changes in response to the forcing. Biologically, this would correspond to \eqref{eq:p}, representing translation, being a rate limiting step.

@@

This would explain the similarity between the sensitivity curves of many parameters - if the parameters in \eqref{eq:s} - \eqref{eq:c} can only act to alter the fixed point values of $p$ and $g$ which the system tends toward, they all play qualitatively identical roles on the time series of $z$. The fact that $\beta m +f_{s}\beta c$ is the only term in \eqref{eq:p} to include $f_{s}$ and $\beta$ may also explain the similarity of the sensitivity curves for these parameters to those found in \eqref{eq:s} - \eqref{eq:c} (Fig. \ref{SensitivityMatrix_scaled}), by the same reasoning. This explanation also suggests that all parameter sets giving the same value of the model's fixed point will give similar model predictions - and there may an enormous space of parameter sets which all give the same fixed point. Fig. \ref{fixedpoint_f} shows a scatterplot of error function value against $\beta m +f_{s}\beta c$ for the 200 estimated parameter sets for the \textit{13\_9} dataset, and demonstrates that, although individual parameter values can be spread across very large ranges, they are correlated in such a way as to give similar model fixed points.

@@

 In contrast to the results of Fig. \ref{InitialResults}, we see relatively tightly estimated parameters, with no estimated values near the initial bounding box. Further, Fig \ref{13_9_f_simplified} shows that this simplified model achieves error values as low as the more complex model.

 We note that \ref{13_9_scatterplot_simplified} shows a strong positive correlation between the minima found for $F$ and $\Theta$. This is consistent with the  normalised sensitivity curves for the \textit{13\_9} dataset, shown in Fig. \ref{Sensitivity_scaled_simplified}. As before, the similarity of the $F$ and $\Theta$ curves suggests that any minima in $F - \Theta$ space will be locally non-unique \cite{Beck}.

@@

 We also note that the parameter estimates come in two distinct clusters.  We further investigate by fixing the value of $\Theta$, and plotting the error as a function of $\mu$ and $F$. Results are shown in Fig. \ref{SimplifiedModelLikelihood}. We find a basin in $ \mu - F$ space, pocketed by many local minima of similar fitness values. The optimisation algorithm will always terminate in one of these local minima, explaining the disconnected clusters shown in Fig.  \ref{13_9_scatterplot_simplified}. Thus the tightness of each peak in Fig. \ref{13_9_hist_simplified} is misleading - the uncertainty in parameter estimates is better reflected by the contours of Fig. \ref{Likelihood_profile_fine}.

@@

In this report, we have presented a system of ODE's to model a recently proposed synthetic RNA regulatory circuit \cite{Rodrigo2012}, and attempted to estimate the model's parameters using existing time series data, with a least squares minimisation approach similar to that found in recent synthetic biology literature \cite{Hu2015}. We have found many of the parameters contained in the full model to be inestimable, and suggested reasons for this. We have further suggested a simplified model, which provides an equally good description of the data with fewer parameters than the initial model.

@@

On the other hand, if a good description of the fluorescence data is what is really of interest, the simplified model may be a better starting point. On the experimental side, $\mu$ and $\Theta$ are both parameters which can be easily estimated independent of our datasets - a realistic range of values could be used to bound the error function profile shown in Fig. \ref{SimplifiedModelLikelihood}. If the experiment is simple to reproduce, another suggestion would be to decrease the gap between timepoints. Further modelling work may focus on improving the modelling of translation, the rate limiting step - \cite{Hu2015}, for example, uses a more detailed model than our single step description of translation. Any additional parameters introduced will likely act on the same timescale as $\mu$, and stand a better chance of being estimable.

@@

\section*{Acknowledgments}
I would like to thank my supervisor, Manish Kushwaha, for his help throughout the project, as well as Alfonso Jaramillo and Brian Munsky for advice on the initial ODE model used, and Shenshi Shen, Guillermo Rodrigo and Boris Kisov for the unpublished data used. In addition I would like to thank James Kermode and Peter Brommer for their helpful discussions on the limitations of the least squares fitting approach, and suggestion of the use of Gaussian Processes as an alternate method - I apologise for not having the time to explore it. Finally I would like the thank Annabelle Ballesta for her advice throughout the project.
% references section
\begin{footnotesize}
\bibliographystyle{./IEEE_bibtex_class/IEEEtran}
\bibliography{./IEEE_bibtex_class/IEEEabrv,library}
\end{footnotesize}
\clearpage
\onecolumn

@@

Figs. \ref{InitialResults_13_9_correlation}, \ref{InitialResults_14_7_correlation} show correlation matrices between parameter sets\footnote{These are computed by viewing each of the estimated parameter sets as a sample from a multivariate random variable. An element of the correlation matrix $R_{ij}$ is then the Pearson correlation coefficient between the $i^\mathrm{th}$ and $j^\mathrm{th}$ random variable - or, in our case, the $i^\mathrm{th}$ and $j^\mathrm{th}$ parameters}. These correlation matrices indicate that there are spaces of parameters within which the fitness function remains approximately constant - for example, in both datasets there exists a positive correlation between values of $\mu$ and $\beta$. Referring to \eqref{eq:p}, this makes heuristic sense - the two parameters may have similar effects on model predictions, and may be able to co-vary in such a way as to leave the model prediction unchanged.
